Eco-evolutionary dynamics of finite populations from first principles

### 2023-04-23 ind_files/03_BD_processes.tex
Rephrasing and restructuring

In appendix \ref{App_density_to_freq}, I use a multivariate version of It\^{o}'s formula to derive a general stochastic equation for the frequencies of each type in the population. Letting $\overline{w} = \sum w_ip_i$ and $\overline{\tau} = \sum \tau_i p_i$ be the average population fitness and turnover respectively, I show in appendix \ref{App_density_to_freq} that the frequency of the $i\textsuperscript{th}$ type in the population $\mathbf{x}(t)$ changes according to the equation:
\begin{equation}
\label{nD_eqn_for_frequencies}
\begin{aligned}
dp_i(t) &= \left[(w_i(\mathbf{x}) - \overline{w})p_i + \mu\left\{Q_i(\mathbf{p}) - p_i\left(\sum\limits_{j=1}^{m}Q_j(\mathbf{p})\right)\right\}\right]dt\\
&- \frac{1}{K}\frac{1}{N_{K}(t)}\left[(\tau_i(\mathbf{x}) - \overline{\tau})p_i + \mu\left\{Q_i(\mathbf{p}) - p_i\left(\sum\limits_{j=1}^{m}Q_j(\mathbf{p})\right)\right\}\right]dt\\
&+ \frac{1}{\sqrt{K}}\frac{1}{N_{K}(t)}\left[\left(A^{+}_{i}(\mathbf{x})\right)^{1/2}dW^{(i)}_t - p_i\sum\limits_{j=1}^{m}\left(A^{+}_{j}(\mathbf{x})\right)^{1/2}dW^{(j)}_t\right]
\end{aligned}
\end{equation}
where $W^{(1)}_t,W^{(2)}_t, \ldots, W^{(m)}_t$ are $m$ independent one-dimensional Wiener processes and I have used the notation $Q_i(\mathbf{p}) = Q_i(\mathbf{x})/N_K(t)$ for notational clarity. I will show below that the first term in this expression describes directional changes in the population composition due to `classical' evolutionary forces such as selection and mutation that occur in deterministic infinite population models. The second term is an additional directional force on population composition that is only seen in finite populations and can be thought of as a biasing `selection' for reduced turnover rate due to an effect similar to gambler's ruin in probability theory. The consequences of this term, as well as connections with previous studies, are discussed in detail in part \ref{part_summary}. Finally, the last term of equation \eqref{nD_eqn_for_frequencies} describes non-directional stochastic effects due to fluctuations and has a `spreading effect' \textcolor{blue}{(Also see equation} \eqref{nD_stoch_RM_noise_term_alt_representation} \textcolor{blue}{in Appendix} \ref{App_alternative_representation} \textcolor{blue}{for a more elegant representation of these noise terms as an integral against a single Wiener process)}. \textcolor{blue}{For the case $m=2, \mu = 0$ (two interacting types of individuals, no mutation), this expression, after using the representation of noise terms presented in Appendix} \ref{App_alternative_representation} \textcolor{blue}{and letting $p = p_1$, becomes the considerably simpler expression
\begin{equation*}
	\begin{aligned}
		dp &= \left[(w_1 - w_2)p(1-p) - \frac{1}{KN_K}(\tau_1-\tau_2)p(1-p)\right]dt\\
		&+ \frac{1}{\sqrt{KN_K}}\left(p(1-p)\left[\tau_1 + (\tau_2 - \tau_1)p\right]\right)^{1/2}dW_t
	\end{aligned}
\end{equation*}
It is easy to see that if we start with an initial condition $p_0 \in [0,1]$, this system will always remain in $[0,1]$, and in that sense, it is `well-behaved'. Well-behavedness at the boundaries for the more general equation} \eqref{nD_eqn_for_frequencies} \textcolor{blue}{can also be checked directly, and has been carried out in Appendix} \ref{App_well_behaved}.

>>>>

In appendix \ref{App_density_to_freq}, I use a multivariate version of It\^{o}'s formula to derive a general stochastic equation for the frequencies of each type in the population. Letting $\overline{w} = \sum w_ip_i$ and $\overline{\tau} = \sum \tau_i p_i$ be the average population fitness and turnover respectively, I show in appendix \ref{App_density_to_freq} that the frequency of the $i\textsuperscript{th}$ type in the population $\mathbf{x}(t)$ changes according to the equation:
\begin{equation}
\label{nD_eqn_for_frequencies}
\begin{aligned}
dp_i(t) &= \left[(w_i(\mathbf{x}) - \overline{w})p_i + \mu\left\{Q_i(\mathbf{p}) - p_i\left(\sum\limits_{j=1}^{m}Q_j(\mathbf{p})\right)\right\}\right]dt\\
&- \frac{1}{K}\frac{1}{N_{K}(t)}\left[(\tau_i(\mathbf{x}) - \overline{\tau})p_i + \mu\left\{Q_i(\mathbf{p}) - p_i\left(\sum\limits_{j=1}^{m}Q_j(\mathbf{p})\right)\right\}\right]dt\\
&+ \frac{1}{\sqrt{K}}\frac{1}{N_{K}(t)}\left[\left(A^{+}_{i}(\mathbf{x})\right)^{1/2}dW^{(i)}_t - p_i\sum\limits_{j=1}^{m}\left(A^{+}_{j}(\mathbf{x})\right)^{1/2}dW^{(j)}_t\right]
\end{aligned}
\end{equation}
where $W^{(1)}_t,W^{(2)}_t, \ldots, W^{(m)}_t$ are $m$ independent one-dimensional Wiener processes and I have used the notation $Q_i(\mathbf{p}) = Q_i(\mathbf{x})/N_K(t)$ for notational clarity. I will show below that the first term in this expression describes directional changes in the population composition due to `classical' evolutionary forces such as selection and mutation that occur in deterministic infinite population models. The second term is an additional directional force on population composition that is only seen in finite populations and can be thought of as a biasing `selection' for reduced turnover rate due to an effect similar to gambler's ruin in probability theory. The consequences of this term, as well as connections with previous studies, are discussed in detail in part \ref{part_summary}. Finally, the last term of equation \eqref{nD_eqn_for_frequencies} describes non-directional stochastic effects due to fluctuations and has a `spreading effect' \textcolor{blue}{(Also see equation} \eqref{nD_stoch_RM_noise_term_alt_representation} \textcolor{blue}{in Appendix} \ref{App_alternative_representation} \textcolor{blue}{for a more elegant representation of these noise terms as an integral against a single Wiener process)}. \textcolor{blue}{For the case $m=2, \mu = 0$ (two interacting types of individuals, no mutation), this expression, after using the representation of noise terms presented in Appendix} \ref{App_alternative_representation} \textcolor{blue}{and letting $p = p_1$, becomes the considerably simpler expression
\begin{equation}
	\label{two_species_stoch_RM}
	\begin{aligned}
		dp &= \left[(w_1 - w_2)p(1-p) - \frac{1}{KN_K}(\tau_1-\tau_2)p(1-p)\right]dt\\
		&+ \frac{1}{\sqrt{KN_K}}\left(p(1-p)\left[\tau_1 + (\tau_2 - \tau_1)p\right]\right)^{1/2}dW_t
	\end{aligned}
\end{equation}
It is easy to see that if we start with an initial condition $p_0 \in [0,1]$, the system defined by equation \eqref{two_species_stoch_RM} will always remain in $[0,1]$, and in that sense, it is `well-behaved'. Similar well-behavedness at the boundaries for the more general equation} \eqref{nD_eqn_for_frequencies} \textcolor{blue}{can also be checked directly, and has been carried out in Appendix} \ref{App_well_behaved}.

### 2023-04-23 ind_files/05_unification_discussion.tex
Rephrasing and restructuring

Each term in equation \eqref{nD_stochastic_Price} lends itself to a simple biological interpretation. The first term, $\textrm{Cov}(w,f)$, is well-understood in the classical Price equation and represents the effect of natural selection in the infinite population setting. In the stochastic Price equation \eqref{nD_stochastic_Price}, the effects of the second term of equation 
\eqref{nD_stochastic_RM} decompose into a selection term $\textrm{Cov}(\tau,f)$ for reduced turnover rates and a transmission bias term that vanishes in the weak mutation ($\mu \to 0$) limit. Following~\cite{constable_demographic_2016} and~\cite{week_white_2021}, I refer to the effect of the covariance term (the second term of equation \eqref{nD_stochastic_Price}) as \emph{noise-induced selection} since it occurs exactly analogously to classical natural selection (but for lower $\tau$) and is induced purely by the finiteness of the population. Since this evolutionary force is unique to finite, fluctuating populations and has been largely overlooked in classical population genetics, it warrants some more detailed discussion. Biologically, the $\textrm{Cov}(\tau,f)$ term (with a negative sign) describes a biasing effect due to differential turnover rates and can intuitively be understood as being similar to gambler's ruin in probability theory through the following reasoning: If a type $i$ has a higher $\tau_i$, it must have a higher birth and death rate and thus experiences more births and deaths in a given time interval than an otherwise equivalent species with a lower $\tau_i$. More events mean greater demographic stochasticity, and types with a higher $\tau_i$ thus tend to be eliminated by a stochastic analog of selection because they experience more chance events (births and deaths) in a given time period and thus have greater stochastic fluctuations. This effect is less visible if the total population size is higher because larger populations generally experience less stochasticity, which is reflected in the $1/N_K$ factor in this term.

This stochastic analog of selection for reduced turnover rates, captured by the second term of equation \eqref{nD_stochastic_RM}, is the force responsible for the `reversal of the direction of deterministic selection' induced by demographic noise in previous studies~\citep{houchmandzadeh_selection_2012, houchmandzadeh_fluctuation_2015, constable_demographic_2016,veller_drift-induced_2017, mcleod_social_2019}. Note that populations composed largely of types that tend to increase the \emph{total} population size $KN_K(t)$ (such as altruists in evolutionary theory and mutualists in ecological communities) will experience a lower magnitude of noise-induced selection compared to populations largely composed of types that do not facilitate such an increase, such as cheaters and highly competitive species. Further, if altruists act by reducing the death rate (rather than increasing the birth rate) of other individuals, their presence causes higher $w$ and lower $\tau$ in the beneficiary individuals, both of which are favored by selection (but note that if they act by increasing the birth rate, they increase the magnitude of negative noise-induced selection disfavoring the beneficiary individual), which explains why this effect preferentially favors mutualists in reversing the direction of deterministic selection in finite population models with fluctuating population sizes if interaction effects are on the death rate~\citep{mcleod_social_2019}. Thus, selection for reduced turnover rate could help explain why cooperation often persists in fluctuating populations in laboratory experiments~\citep{sanchez_feedback_2013} and finite population IbMs of social evolution~\citep{houchmandzadeh_selection_2012,houchmandzadeh_fluctuation_2015,chotibut_evolutionary_2015,behar_fluctuations-induced_2016,mcavoy_public_2018,mcleod_social_2019} despite infinite population models predicting their extinction.

The fact that total population size controls the strength of noise-induced selection also explains why cooperation is favored in the early transient period of population growth~\citep{melbinger_evolutionary_2010} when simulations are initiated from a small population size - In the early transient period, $N_K(t)$ is small, and the biasing effect of differential turnover rates is stronger, thus favoring cooperation. The fact that the entire term scales inversely with the total population size $KN_K(t)$ suggests that the effect of this force is weak for large populations, which explains why the studies that find `reversal of direction of selection' work in settings that allow for conditions such as quasi-neutrality, timescale separation, or a weak selection + weak mutation limit. In all three of these cases, the effects of classical selection vanish, and thus the first term on the RHS of \eqref{nD_stochastic_Price} becomes identically 0. It therefore no longer contributes to the trait frequency dynamics, allowing us to see the (otherwise weak) contributions of the second term.

>>>>

Each term in equation \eqref{nD_stochastic_Price} lends itself to a simple biological interpretation. The first term, $\textrm{Cov}(w,f)$, is well-understood in the classical Price equation and represents the effect of natural selection in the infinite population setting. In the stochastic Price equation \eqref{nD_stochastic_Price}, the effects of the second term of equation \eqref{nD_stochastic_RM} decompose into a selection term $\textrm{Cov}(\tau,f)$ for reduced turnover rates and a transmission bias term that vanishes in the weak mutation ($\mu \to 0$) limit. Following~\cite{constable_demographic_2016} and~\cite{week_white_2021}, I refer to the effect of the covariance term (the second term of equation \eqref{nD_stochastic_Price}) as \emph{noise-induced selection} since it occurs exactly analogously to classical natural selection (but for lower $\tau$) and is induced purely by the finiteness of the population. Biologically, the $\textrm{Cov}(\tau,f)$ term (with a negative sign) describes a biasing effect due to differential turnover rates and can intuitively be understood as being similar to gambler's ruin in probability theory through the following reasoning: If a type $i$ has a higher $\tau_i$, it must have a higher birth and death rate and thus experiences more births and deaths in a given time interval than an otherwise equivalent species with a lower $\tau_i$. More events mean greater demographic stochasticity, and types with a higher $\tau_i$ thus tend to be eliminated by a stochastic analog of selection because they experience more chance events (births and deaths) in a given time period and thus have greater stochastic fluctuations. This effect is less visible if the total population size is higher because larger populations generally experience less stochasticity, which is reflected in the $1/N_K$ factor in this term. \textcolor{blue}{Noise-induced selection for reduced turnover rates has also been interpreted as a selection for `lower reproductive variance'~\citep{gillespie_natural_1974,wang_reproductive_2023}. This latter interpretation is on purely mathematical grounds --- the total turnover $\tau_i x_i$ determines the variance of the distribution obtained as the solution to the SDE for population densities given by equation \eqref{nD_Ito_SDE}.}

\textcolor{red}{\st{Note that types that tend to increase the \emph{total} population size $KN_K(t)$ (such as altruists in evolutionary theory and mutualists in ecological communities) will reduce the magnitude of this effect compared to types that do not facilitate such an increase, such as cheaters and highly competitive species. Further, if altruists/mutualists act by reducing the death rate (rather than increasing the birth rate) of other individuals, their presence causes higher $w$ and lower $\tau$ in the beneficiary individuals, both of which are favored by selection (but note that if they act by increasing the birth rate, they increase the magnitude of negative noise-induced selection disfavouring the beneficiary individual), which explains why this effect preferentially favors mutualists in reversing the direction of deterministic selection in finite population models with fluctuating population sizes if interaction effects are on the death rate~(McLeod and Day, 2019). Thus, selection for reduced turnover rate could help explain why cooperation often persists in fluctuating populations in laboratory experiments and finite population IbMs of social evolution that allow for fluctuating populations despite infinite population models predicting their extinction~(Houchmandzadeh and Vallade, 2012; Sanchez and Gore, 2013; Houchmandzadeh, 2015; Chotibut and Nelson, 2015; Behar et al., 2016; McAvoy et al., 2018; McLeod and Day, 2019)}}.

\textcolor{red}{\st{The fact that total population size controls the strength of noise-induced selection also explains why cooperation is favored in the early transient period of population growth~(Melbinger et al., 2010) when simulations are initiated from a small population size --- In the early transient period, $N_K(t)$ is small, and the biasing effect of differential turnover rates is stronger, thus favoring cooperation. The fact that the entire term scales inversely with the total population size $KN_K(t)$ suggests that the effect of this force is weak for large populations, which explains why the persistence of cooperators is often only observed in restrictive sounding conditions such as quasi-neutrality, timescale separation, or a weak selection + weak mutation limit~(McLeod and Day, 2019). In all three of these cases, the effects of classical selection vanish, and thus the first term on the RHS of }}\eqref{nD_stochastic_Price} \textcolor{red}{\st{becomes identically 0. It therefore no longer contributes to the trait frequency dynamics, allowing us to see the (otherwise weak) contributions of noise-induced selection.}}

### 2023-04-23 ind_files/06_nontechnical_discussion.tex
Rephrasing and restructuring

@@

Chapter \ref{chap_unification} formulates a set of stochastic differential equations derived from our general birth-death formalism which reveal that the results of all these studies can be explained as specific instances of a more general phenomenon we call `noise-induced selection' (following previous studies such as~\cite{constable_demographic_2016} and~\cite{week_white_2021}). In particular, the equations of section \ref{sec_fun_theorems} generically predict a directional evolutionary force --- noise-induced selection --- acting on variation in per-capita turnover rate $\tau$ in the population and favoring types whose per-capita turnover rate is lower than that of the population average. Noise-induced selection is only seen in finite populations, is seen whenever there is differential turnover rate $\tau$ in the system, and arises due to different types of individuals in the population experiencing a different number of stochastic events (birth and death) in a given time interval. This thesis is, to the best of my knowledge, the first time that noise-induced selection has been explicitly formulated and derived in full generality using a general `model-independent' evolutionary framework such as the Price equation (or my generalization thereof). My results thus unify several previous studies focused on specific systems under a single conceptual banner and show that their predictions are a generic outcome in finite, fluctuating populations of any nature as long as there is some heritable variation in per-capita turnover rates.

A recent study showed (among other things) that in a very broad class of competition models of population ecology, finite populations can directionally evolve in a manner opposite to the direction of evolution predicted in the infinite population limit, essentially due to the effects of noise-induced selection~\citep{mazzolini_universality_2022}. My thesis generalizes this particular result of~\citet{mazzolini_universality_2022} to models with arbitrary interaction types and presents the relevant equations in a form that highlights connections with standard results in evolutionary theory such as the Price equation and Fisher's fundamental theorem (see section \ref{sec_fun_theorems}). It is worth noting that the existence of noise-induced selection directly implies (from equation \eqref{nD_stochastic_RM} or \eqref{nD_stochastic_Price}) that evolution is not expected to maximize fitness in finite populations even if fitness is entirely frequency independent as long as there is some (heritable) variation in the turnover rates $\tau_i$, further underscoring the now well-appreciated point that the view of evolution as `climbing a hill' on a fitness landscape and thereby maximizing fitness is rather limited~\citep{grodwohl_theory_2017}.

The equations of section \ref{sec_fun_theorems} also imply that for the evolution of a trait to be \emph{truly} neutral in finite populations (in the sense of all $m$ types in a system having equal fixation/extinction probability if we start with an initial state in which every type has frequency $1/m$), it is not sufficient for the trait in question to be neutral with respect to fitness $w$. Instead, we also require the trait to be neutral with respect to turnover rate $\tau$. Though this is clear from examining equations \eqref{nD_stochastic_RM} or \eqref{nD_stochastic_Price}, I also illustrate a deviation from neutrality explicitly for the $m=2$ case using a simple resource-competition model in Appendix \ref{App_deviation_from_neutrality}. Indeed, in ecological models, previous work in the `quasi-neutral' regime has shown that in finite, fluctuating populations, the equal growth rate of types is not sufficient to ensure equal fixation probabilities and that there is a slight biasing for types with lower turnover rates, sometimes interpreted as a selection `for longevity'~\citep{lin_features_2012, oliveira_advantage_2017,balasekaran_quasi-neutral_2022}. In models of evolutionary game theory in fluctuating finite populations, individuals with lower death rates have higher fixation probability even when growth rates are equalized~\citep{huang_stochastic_2015, czuppon_fixation_2018}. Similarly, models of cell cycle dynamics find that selection favors cell types that periodically arrest their cell cycle relative to non-arresting cells even when their growth rates are equal~\citep{wodarz_effect_2017}. In the language of the birth-death formalism I develop in this thesis, all of these studies equalize the growth rates $w$ of competing types but allow the turnover to vary (by arresting the cell cycle and thus reducing turnover, for example), thus allowing noise-induced selection for reduced turnover to operate in the system.

@@

>>>>

\section{Noise-induced selection is a generic outcome of evolution in finite, fluctuating populations}

@@

Chapter \ref{chap_unification} formulates a set of stochastic differential equations derived from our general birth-death formalism which reveal that the results of all these studies can be explained through a single set of equations. In particular, the equations of section \ref{sec_fun_theorems} generically predict a directional evolutionary force --- noise-induced selection --- acting on variation in per-capita turnover rate $\tau$ in the population and favoring types whose per-capita turnover rate is lower than that of the population average. Noise-induced selection is only seen in finite populations, is seen whenever there is differential turnover rate $\tau$ in the system, and arises due to different types of individuals in the population experiencing a different number of stochastic events (birth and death) in a given time interval. This thesis is, to the best of my knowledge, the first time that noise-induced selection has been explicitly formulated and derived in full generality using a general `model-independent' evolutionary framework such as the Price equation (or my generalization thereof). My results thus unify several previous studies focused on specific systems under a single conceptual banner and show that their predictions are a generic outcome in finite, fluctuating populations of any nature as long as there is some heritable variation in per-capita turnover rates.

\textcolor{blue}{Note that types that tend to increase the \emph{total} population size $KN_K(t)$ (such as altruists in evolutionary theory and mutualists in ecological communities) will reduce the magnitude of this effect compared to types that do not facilitate such an increase, such as cheaters and highly competitive species. Further, if altruists/mutualists act by reducing the death rate (rather than increasing the birth rate) of other individuals, their presence causes higher $w$ and lower $\tau$ in the beneficiary individuals, both of which are favored by selection (but note that if they act by increasing the birth rate, they increase the magnitude of negative noise-induced selection disfavouring the beneficiary individual), which explains why this effect preferentially favors mutualists in reversing the direction of deterministic selection in finite population models with fluctuating population sizes if interaction effects are on the death rate~\citep{mcleod_social_2019}. Thus, selection for reduced turnover rate could help explain why cooperation often persists in fluctuating populations in laboratory experiments and finite population IbMs of social evolution that allow for fluctuating populations despite infinite population models predicting their extinction~\citep{houchmandzadeh_selection_2012,sanchez_feedback_2013,houchmandzadeh_fluctuation_2015,chotibut_evolutionary_2015,behar_fluctuations-induced_2016,mcavoy_public_2018,mcleod_social_2019}}.

\textcolor{blue}{The fact that total population size controls the strength of noise-induced selection also explains why cooperation is favored in the early transient period of population growth~\citep{melbinger_evolutionary_2010} when simulations are initiated from a small population size --- In the early transient period, $N_K(t)$ is small, and the biasing effect of differential turnover rates is stronger, thus favoring cooperation. The fact that the entire term scales inversely with the total population size $KN_K(t)$ suggests that the effect of this force is weak for large populations, which explains why the persistence of cooperators is often only observed in restrictive sounding conditions such as quasi-neutrality, timescale separation, or a weak selection + weak mutation limit~\citep{mcleod_social_2019}. In all three of these cases, the effects of classical selection vanish, and thus the first term on the RHS of }\eqref{nD_stochastic_Price} \textcolor{blue}{becomes identically 0. It therefore no longer contributes to the trait frequency dynamics, allowing us to see the (otherwise weak) contributions of noise-induced selection.}

\section{Noise-induced selection has concrete implications for finite populations}

It is worth noting that the existence of noise-induced selection directly implies (from equation \eqref{nD_stochastic_RM} or \eqref{nD_stochastic_Price}) that evolution is not expected to maximize fitness in finite populations even if fitness is entirely frequency independent as long as there is some (heritable) variation in the turnover rates $\tau_i$, further underscoring the now well-appreciated point that the view of evolution as `climbing a hill' on a fitness landscape and thereby maximizing fitness is rather limited~\citep{grodwohl_theory_2017}. The equations of section \ref{sec_fun_theorems} also imply that for the evolution of a trait to be \emph{truly} neutral in finite populations (in the sense of all $m$ types in a system having equal fixation/extinction probability if we start with an initial state in which every type has frequency $1/m$), it is not sufficient for the trait in question to be neutral with respect to fitness $w$. Instead, we also require the trait to be neutral with respect to turnover rate $\tau$. Though this is clear from examining equations \eqref{nD_stochastic_RM} or \eqref{nD_stochastic_Price}, I also illustrate a deviation from neutrality explicitly for the $m=2$ case using a simple resource-competition model in Appendix \ref{App_deviation_from_neutrality}. Indeed, in ecological models, previous work in the `quasi-neutral' regime has shown that in finite, fluctuating populations, the equal growth rate of types is not sufficient to ensure equal fixation probabilities and that there is a slight biasing for types with lower turnover rates, sometimes interpreted as a selection `for longevity'~\citep{lin_features_2012, oliveira_advantage_2017,balasekaran_quasi-neutral_2022}. In models of evolutionary game theory in fluctuating finite populations, individuals with lower death rates have higher fixation probability even when growth rates are equalized~\citep{huang_stochastic_2015, czuppon_fixation_2018}. Similarly, models of cell cycle dynamics find that selection favors cell types that periodically arrest their cell cycle relative to non-arresting cells even when their growth rates are equal~\citep{wodarz_effect_2017}. In the language of the birth-death formalism I develop in this thesis, all of these studies equalize the growth rates $w$ of competing types but allow the turnover to vary (by arresting the cell cycle and thus reducing turnover, for example), thus allowing noise-induced selection for reduced turnover to operate in the system.

@@

\section{Outlook}

A recent study showed (among other things) that in a very broad class of density-dependent competition models, model-specific details can be ignored in favor of a universal `coarse-grained' description that only includes a small number of quantitities~\citep{mazzolini_universality_2022}. Among the generic predictions of this general description was the presence of noise-induced selection in general competition models. Our work generalizes this particular result of~\cite{mazzolini_universality_2022} to models with arbitrary interaction types and presents the relevant equations in a formalism that is more in accordance with standard biological models such as the Price equation. Our results also show that equations like the replicator-mutator equation and Price equation are asymptotically `universal', in the sense that an arbitrary density-dependent birth-death process with functional forms given by \eqref{nD_functional_forms_for_replicator} will satisfy the replicator-mutator or Price equation as we move to infinitely large populations ($K \to \infty$). Thus, these equations provide coarse-grained `fundamental theorems' that are always satisfied for biological populations that are sufficiently large.
